Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-11-26 patchy-review-responses.tex

%%%%%%%%
\point{}{ 
    A few sections or passages seem out of place:
    \begin{itemize}
        \item[A.]  I think “Simulation methods” should go in the “Materials and meth-
            ods” section, or at least not right between “Establishment of a locally
            adaptive allele due to mutational influx” and “Establishment of a lo-
            cally adaptive allele due to mutational influx”.

        \item[B.] The sentence on p7 beginning “By our definition of fitness. . . ”.

        \item[C.]  The paragraph starting “Also note that. . . ” on p11.
    \end{itemize}
}

\reply
We moved \nameref{ss:simulations} to Materials and Methods,
modified the second sentence,
and rearranged the structure around the third paragraph.

%%%%%%%%
\point{Fig. 4:}{
    In the caption, I think it’s misleading to say that $\migrate$ is ``relatively insensitive to $s_m$''. 
    It’s exponentially sensitive! I don’t understand the blue numbers in the right-hand
    panel. It might be easier to interpret these if they were contour plots
    of the total time to adaptation, colored according to the value of (14),
    although I’m not sure. Also, they aren’t really ``phase diagrams''.
}

\reply
Good points; we meant that it was less sensitive to $s_m$ than we'd otherwise think;
we've replaced this figure with a different one anyhow.

@@

%%%%%%%%
\point{Top of p11 (just above “Multiple patches”):}{should be $\sqrt{2s_m}/\sigma$.
}

\reply
Rephrased. \revref

%%%%%%%%
\point{Between (49) and (51):}{should be ``Green's''.
}

\reply
Always a contentious choice; we're sticking by our position,
and can cite \citet{borodin2002handbook} as backup. ;)

### 2014-11-26 patchy-selection-paper.tex
varous

First we briefly describe the simulations used below for illustration and validation
(and provide the \texttt{R} code used as a supplement).
We simulated forward-time dynamics of the number of alleles of each type in a square grid (either one- or two-dimensional) of demes with fixed size $N$.
Each generation, each individual independently chose to reproduce or not with a probability depending on her type and location in the grid;
locally beneficial alleles were more likely to reproduce.
Each then produced a random, Poisson distributed number of offspring that each, independently, 
either remained in the same location or
migrated a random number of steps in a uniformly chosen cardinal direction;
the probability of migrating $k$ steps was proportional to $2^{-k}$ for $1\le k \le 5$.
\linelabel{rr:sim_details}
Once migrants were distributed, each deme was uniformly resampled back down to $N$ individuals.
(Although we described the simulation in terms of individuals,
we kept track only of total numbers in an equivalent way.)

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles 
far from the original patch where it is already established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this as a branching process,
i.e.\ by assuming that each individual
migrates, then reproduces, independently of all other family members.
Migration takes the individual in a random direction
for a random distance whose mean square is $\sigma^2$,
and if the individual arrives to a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring,
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, the mean number of offspring per individual is $e^{-s_m}\approx 1-s_m$,
and the variance is $\xi^2$.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

@@

This suggests that 
\begin{align} \label{eqn:ratio_K}
  \migrate(x) 
    &= \rho q(x) \times 
    \frac{
        \E[1-(1-p_e)^K]
    }{
        \E[K]/2s_m
    } .
\end{align}
There are now two extremes to consider.
If $\P\{K>1/p_e\}$ is small, then each family has a small chance of establishment,
and the numerator is approximately $p_e \E[K]$.
On the other hand, if $p_e$ is not too small,
then the numerator would be close to 1, as any family that arrives will likely establish.
Combining these, and absorbing the term $\E[W|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:twocase_migrate}
  \migrate(x) &\approx C \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma)  \qquad \text{for } |y| > \sigma/\sqrt{2 s_m} 
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
When we use this below, we set $p_e = 2 s_p / \xi^2$.
Note that the function increases for small $s_m$, 
but decreases if $s_m > |y|^2 / 2 \sigma^2$, 
i.e.\ where the condition holds.

@@

@@

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{(A)} Time to adaptation by new mutation 
      in 3,128 simulations,
      compared to the value predicted by equation X.
      For each of 52 different parameter combinations (Supp Table XX),
      we performed between 40 and 100 independent simulations;
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination (example runs in Supp Figure XX).
      \textbf{(A)} Similar, for time to adaptation by migration
      in 4,158 simulations.
      TODO: put equations on axis labels, describe more, make supp figs.
  }   \label{fig:sim_times}
\end{figure}

@@

Taking these approximations at face value, 
and assuming both rates are small, 
the time until the first of the two patches is colonized by the $B$
allele will be approximately exponentially distributed with mean $1/(2
\mutrate)$.
Following this, the time until the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with mean $1/(\mutrate+\migrate)$.
Both these rates scale linearly with population density ($\rho$) 
and the probability of establishment of a rare allele ($p_e\approx 2 s_b/\xi^2$),
so that the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with population density and beneficial selection coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adaptation, is 
\begin{equation} \label{eqn:parallel_prob}
    \frac{\mutrate}{\mutrate+\migrate(R)} = \frac{ A \mu / (2s_m) }{ A \mu / (2s_m) + \left(R \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2}\exp\left(- \sqrt{2 s_m} R / \sigma \right) },  
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches, at rate $\sqrt{s_m} /\sigma$.

@@

If not yet adapted, a focal patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $R_1, \ldots, R_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}
if patches do not interfere with each other, \linelabel{rr:interference}
is
\begin{equation} \label{eqn:total_migrate}
  \sum_{i=1}^k \migrate(R_i) = C \frac{ 4 \rho s_p s_m }{\xi^2} \sum_{i=1}^{k} \left(R_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m} /\sigma\right).
\end{equation}
If the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}   \label{eqn:migprob}
  \frac{\left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
Equations \eqref{eqn:total_migrate} and \eqref{eqn:migprob} specify the transition rates of the process.
Since the compound parameter $s_p \rho / \xi^2$ is common to both migration and mutation rates,
it functions only as a time scaling of the process, 
and therefore has no effect on the final configuration, namely, 
the sets of patches sharing the genetic basis of their adaptive response.

Also note that we can rescale time by the typical patch size, and introduce a parameter, say $a_k = A_k/\bar A$,
making the properties (other than the time scaling) of the discrete model independent of the \emph{numerical sizes} of the demes themselves.
This is complementary to the results of \cite{softsweepsII}, who showed that multiple mutations are likely to arise \emph{within} a panmictic deme
if the population-scaled mutation rate $2 N \mu$ is greater than $1$.

@@

@@

Intuitively, $h_A(x,y)$ counts each family once if it hits the patch,
while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
Since the expected total time spent in the patch does not depend on the distance at which the family started
if we know that it survived to hit the patch,
it is intuitively reasonable that $u(x,y)$ and $h_A(x,y)$ are simply related.
Following this intuition,
and that $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$,
it turns out (see section~\nameref{ss:hitting_occupation}) that
\begin{align} 
    h_A(x,y) \approx \frac{ u(x,y) }{ \E[K] g(A) } ,
\end{align}
where $g(A)$ is the total expected occupation time of the patch given the family arrives.
Recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \ref{eqn:occupation_integral} and \ref{eqn:migrate_integral} only differ by a factor
that (at least asymptotically) does not depend on the distance between $x$ and $y$,
we can obtain the migration rate by multiplying the equilibrium frequency by this factor.
The probability $f(A)$ that the family does not die out once it arrives at the new patch is
approximately $\E[1-(1-p_e)^K] \approx 2 s_p \E[K] / \xi^2$,
if each member arriving has chance $p_e$ of establishing, independently.
\plr{And if $p_e$ is small?}
% If we let $w$ be the width of the patch, then the remaining term is
% \begin{align}
%   s(w) \simeq \frac{1}{2s_m} \left( 1 - \left( \frac{\pi w \sqrt{2 s_m}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2s_m}/\sigma} \right).
% \end{align}
Combining with the fact that $g(A)\approx 1/(2s_m)$ (see section~\nameref{ss:hitting_occupation}, in particular equation \eqref{eqn:q_migrate_relation}),
we obtain that
\begin{align}
  \label{eqn:migrate} 
  \migrate(y) &\approx q(y) \times \frac{ 4 \rho s_p  s_m }{ \xi^2 }  \\
  &\approx C \frac{ 4 \rho s_p s_m }{ \xi^2 } \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma) .
\end{align}
This does not depend on $A$, because although
$f(A)$ and $g(A)$ in principle depend on the patch size (and geometry),
the dependence is very weak.
For instance, the width of a circular patch only very weakly affects the chance 
of establishment of a new migrant that appears on its edge,
as long as the patch is large enough.
For more discussion, including an expression for the outflux of migrants $\gamma(x)$, see section~\nameref{ss:hitting_occupation}.

@@

>>>>

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles 
far from the original patch where it is already established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this as a branching process,
i.e.\ by assuming that each individual
migrates, then reproduces, independently of all other family members.
Migration takes the individual in a random direction
for a random distance whose mean square is $\sigma^2$,
and if the individual arrives to a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring,
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, 
in this tree the mean number of offspring per individual is $e^{-s_m} < 1$.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

@@

This suggests that 
\begin{align} \label{eqn:ratio_K}
  \migrate(x) 
    &= \rho q(x) \times 
    \frac{
        \E[1-(1-p_e)^K]
    }{
        \E[K]/2s_m
    } .
\end{align}
There are now two extremes to consider.
If $\P\{K>1/p_e\}$ is small, then each family has a small chance of establishment,
and the numerator is approximately $p_e \E[K]$.
On the other hand, if $p_e$ is not too small,
then the numerator would be close to 1, as any family that arrives will likely establish.
Combining these, and absorbing the term $\E[W|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:migrate}
  \migrate(x) &\approx C \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma)  \qquad \text{for } |y| > \sigma/\sqrt{2 s_m} 
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
When we use this below, we set $p_e = 2 s_p / \xi^2$.
Note that the function increases for small $s_m$, 
but decreases if $s_m > |y|^2 / 2 \sigma^2$, 
i.e.\ where the condition holds.
\linelabel{rr:sm_condition}

@@

To test the robustness of our results to the various approximations we used,
we used individual-based simulations on a one-dimensional lattice of 501 demes,
run for 25,000 generations.
More details are given below in \nameref{ss:simulations},
and the number of simulations used and parameter values are given in supplementary tables \ref{stab:mutation_params} and \ref{stab:migration_params}.
To estimate the mean time until adaptation by mutation,
we used one centrally located patch of 99 demes
and a mutation rate of $\mu=10^{-5}$,
and to estimate the mean time until adaptation by migration,
we used two centrally located patches of 99 demes
separated by a varying number of demes,
with one patch initialized to have the $B$ allele at frequency 0.8.
In each case, we measured the ``time to adaptation'' as the amount of time until at least 100 $B$ alleles
were present in the previously unadapted patch.
Figure~\ref{fig:sim_times}
summarizes how the results compare to theory,
excluding parameter combinations where a majority of simulations did not adapt by 25,000 generations
or the patch was not large enough to support a stable population of $B$ alleles.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{(A)} 
      Time until adaptation by mutation in an isolated patch for different parameter combinations,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (as $1/\mutrate$, horizontal axis).
      \textbf{(B)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate} (as $1/\migrate$, horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:migration_params} and \ref{stab:mutation_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
  }   \label{fig:sim_times}
\end{figure}

@@

@@

Taking these approximations at face value, 
and assuming both rates are small, 
the time until the first of the two patches is colonized by the $B$
allele will be approximately exponentially distributed with mean $1/(2
\mutrate)$.
Following this, the time until the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with mean $1/(\mutrate+\migrate)$.
Both these rates scale linearly with population density ($\rho$) 
and the probability of establishment of a rare allele ($p_e\approx 2 s_b/\xi^2$),
so that the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with population density and beneficial selection coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adaptation, is 
\begin{equation} \label{eqn:parallel_prob}
    \frac{\mutrate}{\mutrate+\migrate(R)} = \frac{ A \mu / (2s_m) }{ A \mu / (2s_m) + \left(R \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2}\exp\left(- \sqrt{2 s_m} R / \sigma \right) },  
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches, on the spatial scale $\sigma/\sqrt{s_m}$. \revpoint{2}{13}

@@

If not yet adapted, a focal patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $R_1, \ldots, R_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}
if patches do not interfere with each other, \linelabel{rr:interference}
is
\begin{equation} \label{eqn:total_migrate}
  \sum_{i=1}^k \migrate(R_i) = C \frac{ 4 \rho s_p s_m }{\xi^2} \sum_{i=1}^{k} \left(R_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m} /\sigma\right).
\end{equation}
If the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}   \label{eqn:migprob}
  \frac{\left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
Equations \eqref{eqn:total_migrate} and \eqref{eqn:migprob} specify the transition rates of the process.

Since the compound parameter $s_p \rho / \xi^2$ is common to both migration and mutation rates,
it functions only as a time scaling of the process, 
and therefore has no effect on the final configuration, namely, 
the sets of patches sharing the genetic basis of their adaptive response.
Also note that we can rescale time by the typical patch size, and introduce a parameter, say $a_k = A_k/\bar A$,
making the properties (other than the time scaling) of the discrete model independent of the \emph{numerical sizes} of the demes themselves.
This is complementary to the results of \cite{softsweepsII}, who showed that multiple mutations are likely to arise \emph{within} a panmictic deme
if the population-scaled mutation rate $2 N \mu$ is greater than $1$.

@@

First we briefly describe the simulations we used for illustration and validation
(the \texttt{R} code used is provided as a supplement).
We simulated forward-time dynamics of the number of alleles of each type in a square grid (either one- or two-dimensional) of demes with fixed size $N$.
Each generation, each individual independently chose to reproduce or not with a probability depending on her type and location in the grid;
locally beneficial alleles were more likely to reproduce.
Each then produced a random, Poisson distributed number of offspring that each, independently, 
either remained in the same location with probabilty $0.8$ or
migrated a random number of steps in a uniformly chosen cardinal direction;
the probability of migrating $k$ steps was proportional to $2^{-k}$ for $1\le k \le 5$.
\linelabel{rr:sim_details}
Once migrants were distributed, each deme was uniformly resampled back down to $N$ individuals.
(Although we described the simulation in terms of individuals,
we kept track only of total numbers in an equivalent way.)

@@

Intuitively, $h_A(x,y)$ counts each family once if it hits the patch,
while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
Since the expected total time spent in the patch does not depend on the distance at which the family started
if we know that it survived to hit the patch,
it is intuitively reasonable that $u(x,y)$ and $h_A(x,y)$ are simply related.
Following this intuition,
and that $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$,
it turns out (see section~\nameref{ss:hitting_occupation}) that
\begin{align} 
    h_A(x,y) \approx \frac{ u(x,y) }{ \E[K] g(A) } ,
\end{align}
where $g(A)$ is the total expected occupation time of the patch given the family arrives.
Recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \ref{eqn:occupation_integral} and \ref{eqn:migrate_integral} only differ by a factor
that (at least asymptotically) does not depend on the distance between $x$ and $y$,
we can obtain the migration rate by multiplying the equilibrium frequency by this factor.
The probability $f(A)$ that the family does not die out once it arrives at the new patch is
approximately $\E[1-(1-p_e)^K] \approx 2 s_p \E[K] / \xi^2$,
if each member arriving has chance $p_e$ of establishing, independently.
\plr{And if $p_e$ is small?}
% If we let $w$ be the width of the patch, then the remaining term is
% \begin{align}
%   s(w) \simeq \frac{1}{2s_m} \left( 1 - \left( \frac{\pi w \sqrt{2 s_m}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2s_m}/\sigma} \right).
% \end{align}
Combining with the fact that $g(A)\approx 1/(2s_m)$ (see section~\nameref{ss:hitting_occupation}, in particular equation \eqref{eqn:q_migrate_relation}),
we obtain that XXX FIX
\begin{align}
  \label{eqn:orig_migrate} 
  \migrate(y) &\approx q(y) \times \frac{ 4 \rho s_p  s_m }{ \xi^2 }  \\
  &\approx C \frac{ 4 \rho s_p s_m }{ \xi^2 } \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma) .
\end{align}
This does not depend on $A$, because although
$f(A)$ and $g(A)$ in principle depend on the patch size (and geometry),
the dependence is very weak.
For instance, the width of a circular patch only very weakly affects the chance 
of establishment of a new migrant that appears on its edge,
as long as the patch is large enough.
For more discussion, including an expression for the outflux of migrants $\gamma(x)$, see section~\nameref{ss:hitting_occupation}.

@@

%%%%%%%
\section{Parameter values for simulations}
\label{apx:parameter_tables}

\input{mutation-parameters-table.tex}

\input{migration-parameters-table.tex}
